A Type System for Safe and Scalable Parallelism using Logical Regions

### 2012-07-10 doc/popl2013/example.tex
Major work on the example and some edits to the intro

@@

\begin{lstlisting}[float={t},label={lst:circuit_ex},caption={Circuilt Simulation}]
-- <voltage,current,charge,capacitance>
type CircuitNode        = <float,float,float,float>
-- < owned node, owned or ghost node, resistance, current>
type CircuitWire<rn,rg>  = <CircuitNode@rn, CircuitNode@(rn,rg),float,float>

type node_list<rl,rn>       = < CircuitNode@rn, node_list<rl,rn>@rl >
type wire_list<rl,rw,rn,rg>= < CircuitWire<rn,rg>@rw, wire_list<rl,rw,rn,rg>@rl >

type CircuitPiece<rl,rw,rn> = rr[rpw,rpn,rg]
                            < wire_list<rl,rpw,rpn,rg>@rl, node_list<rl,rpn>@rl >         
                            where rpn <= rn and rg <= rn and rpw <= rw and
                                  rpn * rg and rn * rw and rl * rn and rl * rw

-- Simulation initialization and invocation
function simulate_circuit[rl,rw,rn] ( all_nodes : node_list<rl,rn>@rl, 
                            all_wires : wire_list<rl,rw,rn,rn>@rl, steps : int ), 
      reads(rn,rw,rl), writes(rn,rw,rl) : bool = 
  let pc : <coloring(rn),multicoloring(rn),coloring(rw)> 
            = color_circuit[rn,rw,rl](all_nodes,all_wires) in
  -- Disjoint partition for the owned nodes of each piece
  partition rn using pc.1 as rn0,rn1 in
  -- Aliased partition for ghost nodes of each piece
  partition rn using pc.2 as rg0,rg1 in
  -- Disjoint partition for the owned wires of each piece
  partition rw using pc.3 as rw0,rw1 in
  let lists0 : <wire_list<rl,rw0,rn0,rg0>@rl,node_list<rl,rn0>@rl> = 
        build_lists[rl,rw,rn,rw0,rn0,rg0](all_nodes,all_wires,pc.1,pc.2,pc.3,0) in
  let piece0 : CircuitPiece<rl,rw,rn> = 
        pack lists0 as CircuitPiece<rl,rw,rn>[rw0,rn0,rg0] in
  let lists1 : <wire_list<rl,rw1,rn1,rg1>@rl,node_list<rl,rn1>@rl> =
        build_lists[rl,rw,rn,rw1,rn1,rg1](all_nodes,all_wires,pc.1,pc.2,pc.3,1) in
  let piece1 : CircuitPiece<rl,rw,rn> = 
        pack lists1 as CircuitPiece<rl,rw,rn>[rw1,rn1,rg1] in
      execute_time_steps[rl,rw,rn](piece0,piece1,steps)

-- Time Step Loop
function execute_time_steps[rl,rw,rn] ( p0 : CircuitPiece<rl,rw,rn>, 
      p1 : CircuitPiece<rl,rw,rn>, steps : int ) , reads(rn,rw,rl), writes(rn,rw) : bool = 
  if steps #$<$# 1 then true else
  unpack p0 as piece0 : CircuitPiece<rl,rw,rn>[rw0,rn0,rg0] in 
  unpack p1 as piece1 : CircuitPiece<rl,rw,rn>[rw1,rn1,rg1] in
  let _ : bool = calc_new_currents[rl,rw0,rn0,rg0](piece0.1) in
  let _ : bool = calc_new_currents[rl,rw1,rn1,rg1](piece1.1) in
  let _ : bool = distribute_charge[rl,rw0,rn0,rg0](piece0.1) in
  let _ : bool = distribute_charge[rl,rw1,rn1,rg1](piece1.1) in
  let _ : bool = update_voltage[rl,rn0](piece0.2) in
  let _ : bool = update_voltage[rl,rn1](piece1.2) in
      execute_time_steps[rl,rw,rn](p0,p1,steps-1)

function color_circuit[rn,rw,rl] ( all_nodes : node_list<rl,rn>@rl, 
                               all_wires : wire_list<rl,rw,rn>@rl ), 
        reads(rn,rw,rl) : <coloring(rn), multicoloring(rn), coloring(rw)> =  
  -- Invoke programmer chosen coloring algorithm (e.g. METIS)
  -- return owned, ghost, wire colorings

-- Helper method
function build_lists[rl,rw,rn,rpw,rpn,rg] ( nodes : node_list<rl,rn>@rl, 
       wires : wire_list<rl,rw,rn>@rl, oc : coloring(rn), gc : multicoloring(rn), 
       wc : coloring(rw), c : int), reads(rn,rw,rl), writes(rl) 
       : < wire_list<rl,rpw,rpn,rg>@rl, node_list<rl,rpn>@rl > = 
  -- Construct lists of node and wire pointers for the given colorings
\end{lstlisting}

\begin{lstlisting}[float={t},label={lst:circuit_leaf},caption={Circuilt Leaf Tasks}]
function calc_new_currents[rl,rw,rn,rg] ( ptr_list : wire_list<rl,rw,rn,rg>@rl ), 
      reads(rl,rw,rn,rg), writes(rw) : bool =
  if isnull(ptr_list) then true else
  let wire_node : wire_list<rl,rw,rn,rg> = read(ptr_list) in
  let wire : CircuitWire<rn,rg> = read(wire_node.1) in
  let in_node : CircuitNode = read(wire.1) in
  let out_node : CircuitNode = read(wire.2) in
  let current : float = (in_node.1 - out_node.1) /  wire.3 in 
  let new_wire : CircuitWire<rn,rg> = <wire.1,wire.2,wire.3,current> in
  let _ : CircuitWire<rn,rg>@rw = write(wire_node.1, new_wire) in
      calc_new_currents[rl,rw,rn,rg](wire_node.2)

>>>>

Listing~\ref{lst:circuit_ex} shows code for a
circuit simulation written in the core Legion language introduced in
Section~\ref{sec:legioncore} (with one exception described below).  The
circuit simulation takes as input a large graph of circuit elements and wires
connecting the elements.  To perform the simulation in parallel the graph
is partitioned into pieces.  After partitioning
the graph the simulation is run for a number of time steps.  Each time step
consists of three phases: calculating the current on each wire, distributing
charge to the nodes to which each wire is connected, and updating the voltage
of each node.

The top level function in the application is {\tt simulate\_circuit} (line 15).  The {\tt simulate\_circuit} 
function is parameterized on the regions {\tt rl}, {\tt rw}, and {\tt rn} (line 15) which 
implies that the when the function is invoked it must have privileges for the regions 
bound to those parameters.  Furthermore, these privileges must be for both reading 
and writing these regions (line 17).  The only logical regions which {\tt simulate\_circuit} will 
be allowed to access are these regions and their logical subregions.  This property will be 
enforced by the type system introduced in Section~\ref{sec:legioncore}.  {\tt simulate\_circuit}
takes pointers to the lists of nodes and wires that describe the graph.  The nodes
are stored in the {\tt rn} region while wires are stored in the {\tt rw} region.  The
lists live in the {\tt rl} region.

The first step in {\tt simulate\_circuit} is to partition the graph into pieces.  Partitioning
is done by constructing {\em colorings} using the {\tt color\_circuit} function (line 18-19).  A
coloring is a mapping from region elements to colors with each color corresponding to a new subregion
to be created.  The algorithm for performing a coloring is an application specific decision made by 
the programmer.  Note, that in the case of circuit, it would be impossible to
pick a single static partitioning scheme that would do well for all graphs.  Dynamic partitioning
provides the programmer the flexibility to make partitioning decisions based on input data.

The {\tt color\_circuit} function returns three colorings.  The first coloring is a total 
coloring that maps each node in the graph the piece that owns it.  The third coloring is a total
coloring that maps each wire in the graph to a piece which owns on of its nodes.  The second coloring captures a
different property of the graph: those nodes which must be shared between different pieces of the
graph to perform computations in parallel.  These are often referred to as {\em ghost nodes}.  We
describe how ghost node regions are used in conjunction with the owned nodes later.
Note that its possible some ghost nodes are shared between multiple pieces, and can be colored multiple ways.  Thus,
the second coloring is referred to as a {\em multicoloring} because it might map the same element
to multiple colors.

After creating the colorings, the application partitions the region into {\em subregions} based on 
the colorings (lines 21-25).  For simplicity, our example only partitions the circuit into two pieces.  Lines
21 and 23 create two different partitions with two different subregions of the {\tt rn} region.  The first partition is
{\em disjoint} because it is based on a coloring while the second partition is {\em aliased}
because it is based on a multicoloring which does not imply disjointness of subregions.  While 
aliased partitions based on multicolorings are not a part of core Legion described in Section~\ref{sec:legioncore}, 
they are functionally equivalent to creating single partitions for each color in the multicoloring.

Disjoint partitions introduce
disjointness constraints into the static environment indicating that all the subregions in the partition are 
disjoint (e.g. {\tt rn0}$*${\tt rn1} on line 21 says {\tt rn0} is disjoint from {\tt rn1}).
Both disjoint and aliased partitions introduce constraints into the environment indicating that all
created regions are logical subregions of the partitioned parent region (e.g. {\tt rn0}$<${\tt rn}).

The circuit example illustrates an example of partitioning called {\em allocate-then-partition}
where the data structure is allocated first and parititoned later.  Alternatively, a
{\em partition-then-allocate} scheme would allow the creation and paritioning of regions that
are later populated with data.  Sequoia supports a static form of the former while DPJ supports
a static form of the later.  Legion enables both approaches dynamically.

After partitioning the circuit into pieces, the application creates instances of {\tt CircuitPiece} 
(defined on lines 9-12) for each piece (lines 28-33).  {\tt CircuitPiece} is a bounded existential 
type called a {\em region relationship}.  
Region relationships allow the programmer to {\em pack} a group of regions together and remember properties about
them such as disjointness and sub-region relationships (lines 29 and 33).  The type system verifies 
the properties hold statically when packing so when region relationships are {\em unpacked} the same properties can
be reintroduced into the static checking environment (lines 40-41).  One important feature of Legion
is that privileges cannot be packed in a region relationship but can only be passed through function calls.

The {\tt execute\_time\_steps} function represents the primary loop of the simulation (lines 37-48).  Here
we see the importance of having multiple views onto the same logical region.  Both the the 
{\tt calc\_new\_currents} and {\tt distribute\_charge} function calls bind the owned region of a piece 
and the ghost region of a piece.  In the case of {\tt calc\_new\_currents} these regions only need
read privileges, allowing both instances of {\tt calc\_new\_currents} to be run in parallel.  In the
case of {\tt distribute\_charge} the privilege is for a reduction which can also be done 
in parallel because of the atomic and commutative nature of reductions.  If it were only possible to have a single
partition of the region {\tt rn} it would have been impossible to describe these data sharing patterns.

Listing~\ref{lst:circuit_leaf} shows the leaf functions for each phase of the computation.  Each
function iterates over the list of wires or nodes for its piece.  Each function specifies the privileges
that it must have for all of its regions (lines 2,14,23).  These privileges are statically checked
to match the operations performed inside of each function (e.g. read and write).  In the case of {\em reduce}
the privilege must also specify which reduction function is being used (line 14).

In addition to privileges, functions can also specify {\em coherence} on regions.  Coherence specifies what 
values the function is allowed to see from other functions using the same region.  If not otherwise specified,
coherence defaults to {\em exclusive} which means the function must appear to execute as if the whole program
were run sequentialy, which we refer to as {\em program order}.  Line 14 shows an example of a relaxed
coherence mode called {\tt atomic} which says that reductions to {\tt rn} and {\tt rg} must appear atomic relative to other
functions using those regions.  The most relaxed coherence mode is {\tt simult} which means multiple functions
all with simultaneous coherence are allowed to access the region concurrently.

@@

\begin{lstlisting}[float={t},label={lst:circuit_leaf},caption={Circuilt Leaf Functions}]
function calc_new_currents[rl,rw,rn,rg] ( ptr_list : wire_list<rl,rw,rn,rg>@rl ), 
      reads(rl,rw,rn,rg), writes(rw) : bool =
  if isnull(ptr_list) then true else
  let wire_node : wire_list<rl,rw,rn,rg> = read(ptr_list) in
  let wire : CircuitWire<rn,rg> = read(wire_node.1) in
  let in_node : CircuitNode = read(wire.1) in
  let out_node : CircuitNode = read(wire.2) in
  let current : float = (in_node.1 - out_node.1) /  wire.3 in 
  let new_wire : CircuitWire<rn,rg> = <wire.1,wire.2,wire.3,current> in
  let _ : CircuitWire<rn,rg>@rw = write(wire_node.1, new_wire) in
      calc_new_currents[rl,rw,rn,rg](wire_node.2)

### 2012-07-10 doc/popl2013/intro.tex
Major work on the example and some edits to the intro

% This is a description of how the listings should be formatted.
% It can go anywhere before the listings.
\lstset{
  captionpos=b,
  language=Haskell,
  basicstyle=\scriptsize,
  numbers=left,
  numberstyle=\tiny,
  columns=fullflexible,
  stepnumber=1,
  escapechar=\#,
  keepspaces=true,
  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
  morekeywords={function,rr,int,float,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,multicoloring,color,newcolor,atomic,simultaneous},
  deletekeywords={float,head,min,max}
}

\begin{lstlisting}[float={t},label={lst:circuit_ex},caption={Circuilt Simulation}]
-- <voltage,current,charge,capacitance>
type CircuitNode        = <float,float,float,float>
-- < owned node, owned or ghost node, resistance, current>
type CircuitWire<rn,rg>  = <CircuitNode@rn, CircuitNode@(rn,rg),float,float>

@@

-- Helper methods
function color_circuit[rn,rw,rl] ( all_nodes : node_list<rl,rn>@rl, 
                               all_wires : wire_list<rl,rw,rn>@rl ), 
        reads(rn,rw,rl) : <coloring(rn), multicoloring(rn), coloring(rw)> =  
  -- Invoke programmer chosen coloring algorithm (e.g. METIS)
  -- return owned, ghost, wire colorings

function build_lists[rl,rw,rn,rpw,rpn,rg] ( nodes : node_list<rl,rn>@rl, 
       wires : wire_list<rl,rw,rn>@rl, oc : coloring(rn), gc : multicoloring(rn), 
       wc : coloring(rw), c : int), reads(rn,rw,rl), writes(rl) 
       : < wire_list<rl,rpw,rpn,rg>@rl, node_list<rl,rpn>@rl > = 
  -- Construct lists of node and wire pointers for the given colorings
\end{lstlisting}

Legion programs organize data into a forest of {\em logical regions}.  Logical
regions express {\em locality} (data that will be used together, and therefore colocated if possible in the same
physical memory) and {\em independence} (disjoint data that may be operated on in parallel).
Researchers have previously explored language designs with hierarchical decomposition of data to express
locality and independence for high performance programming. Two recent examples, Sequoia \cite{Fatahalion06} and 
Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a mechanism to partition the heap recursively
into a tree of collections of data.  The two designs are different in many aspects, but agree that there is
a single tree-shaped partitioning of data that can be checked fully statically (see Section~\ref{sec:related} for more
discussion of related work).

\begin{lstlisting}[float={t},label={lst:circuit_leaf},caption={Circuilt Leaf Tasks}]
function calc_new_currents[rl,rw,rn,rg] ( ptr_list : wire_list<rl,rw,rn,rg>@rl ), 
      reads(rl,rw,rn,rg), writes(rw) : bool =
  if isnull(ptr_list) then true else
  let wire_node : wire_list<rl,rw,rn,rg> = read(ptr_list) in
  let wire : CircuitWire<rn,rg> = read(wire_node.1) in
  let in_node : CircuitNode = read(wire.1) in
  let out_node : CircuitNode = read(wire.2) in
  let current : float = (in_node.1 - out_node.1) /  wire.3 in 
  let new_wire : CircuitWire<rn,rg> = <wire.1,wire.2,wire.3,current> in
  let _ : CircuitWire<rn,rg>@rw = write(wire_node.1, new_wire) in
      calc_new_currents[rl,rw,rn,rg](wire_node.2)

function distribute_charge[rl,rw,rn,rg] ( ptr_list : wire_list<rl,rw,rn,rg>@rl ), 
      reads(rl,rw,rn), reduces(reduce_charge,rn,rg), atomic(rn,rg) : bool =
  if isnull(ptr_list) then true else
  let wire_node : wire_list<rl,rw,rn,rg> = read(ptr_list) in
  let wire : CircuitWire<rn,rg> = read(wire_node.1) in
  let _ : CircuitNode@rn = reduce(reduce_charge, wire.1, wire.4) in
  let _ : CircuitNode@(rn,rg) = reduce(reduce_charge, wire.2, wire.4) in
      distribute_charge[rl,rw,rn,rg](wire_node.2)

function update_voltage[rl,rn] ( ptr_list : node_list<rl,rn>@rl ), 
      reads(rl,rn), writes(rn) : bool = 
  if isnull(ptr_list) then true else
  let node_node : node_list<rl,rn> = read(ptr_list) in
  let node : CircuitNode = read(node_node.1) in
  let voltage : float = (node.3/node.4) in
  let new_node : CircuitNode = <voltage,node.2,node.3,node.4> in
  let _ : CircuitNode@rn = write(node_node.1, new_node) in
      update_voltage[rl,rn](node_node.2)

-- Reduction function for distribute charge
function reduce_charge ( node : CircuitNode, current : float ) : CircuitNode =
    let new_charge : float = node.3 + current in
        < node.1,new_charge,node.3,node.4>
\end{lstlisting}

Our own experience with Sequoia and many high-performance applications
written in the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that for many practical situations a
single, fairly static partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions need to be dynamically computed.
Furthermore, it is useful to allow mutliple partitions simultaneously of
the same data, providing different views on to that data. Thus, in Legion the 
handling of regions and partitioning of regions is much more dynamic than in 
previous designs.  In particular,
\begin{itemize}
\item  logical regions are first-class values in Legion
and may be dynamically allocated and stored in data structures;

\item  a logical region may be dynamically partitioned in multiple different ways.
\end{itemize}
A consequence of these decisions is that, through partitioning, a
given datum may belong to multiple different regions.  For example, if
a region $R$ participates in three separate disjoint partition
operations and each partition allocates every element of $R$ to a
subregion (partitions need not be total in Legion), then every element
of $R$ is included in three different subregions.  This is why we use
the term {\em logical} regions: language-level regions serve to name
sets of data but do not imply anything about physical layout.  There
is a separate system of {\em physical regions} at runtime that hold
copies of the data in the language-level logical regions.  In fact,
the Legion implementation routinely maintains multiple distinct
physical regions, each with a copy of the data in the corresponding
logical region, for performance reasons.

A further consequence of placing the same data in multiple logical
regions as a result of runtime computation is that Legion has a very
hard static alias analysis problem on regions, which poses a problem
for parallel computation because knowing regions do not alias is the
basis for safe parallel execution.  This is why languages such as
Sequoia and DPJ allow only a single static partition of each
region---it simplifies static analysis of regions to the point that it
is practical.

Legion's execution model is that by default, a program's semantics is
its meaning as a sequential program, which we refer to as the {\em
program order} execution.  While the
implementation is not our focus here, if two functions use
disjoint data (because all of the regions they use are disjoint), then
Legion may execute them simultaneously.  Also like DPJ,
Legion has a system of {\em permissions} ({\em read}, {\em write}, and
{\em reduce}) on regions and an orthogonal system of region {\em
coherence} modes ({\em excl}, {\em atomic}, and {\em simultaneous})
that increase the number of situations in which functions can be
executed in parallel.

The important difference bewteen Legion and the more static approaches
is that the hardest part of the tests to justify parallel execution is
done dynamically instead of statically.  The key observation is that
alias analysis is hard statically but very easy and relatively cheap
dynamically.  Thus, Legion falls between fully static systems, such as
Sequoia and DPJ, and fully dynamic approaches such as transactional
memory.  Legion still has a significant static component; in
particular, the safety of permissions is checked statically and region
pointer dereferences are checked statically.  The dynamic alias tests
are done at the granularity of regions and one check on a region
itself often serves to prove the safety of many subsequent region
accesses.  In contrast, because transactional memory has no mechanism
for grouping heap data into larger collections, it much test the
overlap between two sets of data on each individual element, which is
much more expensive.

>>>>

Legion programs organize data into a forest of {\em logical regions}.  Logical
regions express {\em locality} (data that will be used together, and therefore colocated if possible in the same
physical memory) and {\em independence} (disjoint data that may be operated on in parallel).
Researchers have previously explored language designs with hierarchical decomposition of data to express
locality and independence for high performance programming. Two recent examples, Sequoia \cite{Fatahalian06} and 
Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a mechanism to partition the heap recursively
into a tree of collections of data.  The two designs are different in many aspects, but agree that there is
a single tree-shaped partitioning of data that can be fully checked statically (see Section~\ref{sec:related} for more
discussion of related work).

Our own experience with writing high-performance applications in both 
Sequoia and the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that for many practical situations a
single, fairly static partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions need to be dynamically computed.
Furthermore, it is useful to allow mutliple partitions simultaneously of
the same data, providing different views on to that data. Thus, in Legion the 
handling of regions and partitioning of regions is much more dynamic than in 
previous designs.  In particular,
\begin{itemize}
\item  logical regions are first-class values in Legion
and may be dynamically allocated and stored in data structures;

\item  a logical region may be dynamically partitioned in multiple different ways.
\end{itemize}
A consequence of these decisions is that, through partitioning, a
given datum may belong to multiple different regions.  For example, if
a region $R$ participates in three separate disjoint partition
operations and each partition allocates every element of $R$ to a
subregion (partitions need not be total in Legion), then every element
of $R$ is included in three different subregions.  This is why we use
the term {\em logical} regions: language-level regions serve to name
sets of data but do not imply anything about physical layout or placement
in the memory hierarchy.  There
is a separate system of {\em physical regions} at runtime that hold
copies of the data in the language-level logical regions.  In fact,
the Legion implementation routinely maintains multiple distinct
physical regions, each with a copy of the data in the corresponding
logical region, for performance reasons.

A further consequence of placing the same data in multiple logical
regions is that it introduces aliasing which makes it difficult 
to statically figure out which computations can be run in parallel.  
One approach is to outlaw aliasing which allows for a static analysis
to discover all available parallelism and schedule it entirely at
compile-time, avoiding any runtime overhead.  Languages such as
Sequoia and DPJ take this approach and allow only a single static partition of any region 
to eliminate region aliasing at the cost of the flexibility of their language.

% Putting this here so that we can get the code in a good place
\lstset{
  captionpos=b,
  language=Haskell,
  basicstyle=\scriptsize,
  numbers=left,
  numberstyle=\tiny,
  columns=fullflexible,
  stepnumber=1,
  escapechar=\#,
  keepspaces=true,
  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
  morekeywords={function,rr,int,float,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,multicoloring,color,newcolor,atomic,simultaneous},
  deletekeywords={float,head,min,max}
}
\begin{lstlisting}[float={t},label={lst:circuit_ex},caption={Circuilt Simulation}]
--                        <voltage,current,charge,capacitance>
type CircuitNode        = <float,float,float,float>
--                     < owned node, owned or ghost node, resistance, current>
type CircuitWire<rn,rg>  = <CircuitNode@rn, CircuitNode@(rn,rg),float,float>

@@

function color_circuit[rn,rw,rl] ( all_nodes : node_list<rl,rn>@rl, 
                               all_wires : wire_list<rl,rw,rn>@rl ), 
        reads(rn,rw,rl) : <coloring(rn), multicoloring(rn), coloring(rw)> =  
  -- Invoke programmer chosen coloring algorithm (e.g. METIS)
  -- return owned, ghost, wire colorings

-- Helper method
function build_lists[rl,rw,rn,rpw,rpn,rg] ( nodes : node_list<rl,rn>@rl, 
       wires : wire_list<rl,rw,rn>@rl, oc : coloring(rn), gc : multicoloring(rn), 
       wc : coloring(rw), c : int), reads(rn,rw,rl), writes(rl) 
       : < wire_list<rl,rpw,rpn,rg>@rl, node_list<rl,rpn>@rl > = 
  -- Construct lists of node and wire pointers for the given colorings
\end{lstlisting}

% Moving this to the circuit example section since it fits better there
% and some of this is redundant
%Legion's execution model is that by default, a program's semantics is
%its meaning as a sequential program, which we refer to as the {\em
%program order} execution.  While the
%implementation is not our focus here, if two functions use
%disjoint data (because all of the regions they use are disjoint), then
%Legion may execute them simultaneously.  Also like DPJ,
%Legion has a system of {\em permissions} ({\em read}, {\em write}, and
%{\em reduce}) on regions and an orthogonal system of region {\em
%coherence} modes ({\em excl}, {\em atomic}, and {\em simultaneous})
%that increase the number of situations in which functions can be
%executed in parallel.

The important difference bewteen Legion and the more static approaches
is that Legion allows aliasing, but does so in a structured way that minimizes
runtime overheads.  The key observation is that
alias analysis is hard statically but very easy dynamically and is relatively cheap
when done at the granularity of regions instead of individual memory
locations.  Thus, Legion falls between fully static systems, such as
Sequoia and DPJ, and fully dynamic approaches such as transactional
memory.  Legion still has a significant static component; in
particular, the safety of permissions is checked statically and region
pointer dereferences are checked statically.  The dynamic alias tests
for independence are done at the granularity of regions and one check on a region
itself often serves to prove the independence of many subsequent region
uses.  In contrast, because transactional memory has no mechanism
for grouping heap data into larger collections, it must test the
overlap between two sets of data on each individual element, which is
much more expensive.
